---
title: "Survival Analysis Demo"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true          
    social: menu
---

```{r setup, include=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ezplot)
library(survival)
library(flexdashboard)

# load data
load("sysdata.rda")

# set color-blind friendly colors
red = "#D55E00"
blue = "#0072B2"
green = "#009E73"
yellow = "#F0E442"
purple = "#CC79A7"
```

### About

Radiotherapy has dramatically improved prognosis for patients with Hodgkin 
disease (HD) in the last several decades. Unfortunately, its carcinogenic nature 
has caused a significant increase in the risk of secondary malignancies in HD survivors. Secondary cancers are the leading cause of death among 15-year 
HD survivors. Breast cancer is the most common solid tumor diagnosed in female HD survivors and the risk varies with patient and treatment-related factors. 

The dataset used in this demo contains breast cancer incidence in HD female 
survivors, as well as age at treatment, site treated, ovarian function, 
radiotherapy dose, and other treatment modalities.

The goal is to investigate the time to the onset of breast cancer in patients 
with HD treated with and without radiotherapy. In search for an adequate model 
for the data, we tried a number of survival models, namely, the non-parametric
Kaplan Meier estimates, the classic cox regression, the proportional odds model 
and the PHPH cure model, investigating difference in breast cancer specific 
survival probabilities between patients who received radiotherapy and patients 
who did not.

Made by [Cabaceo LLC](https://cabaceo.com).


### Time to breast cancer

```{r}
plt = mk_boxplot(dat)
p = plt("RT", "TTOBRC", notched = T)
add_labs(p, ylab = "Months",
         title = "Distribution of months to breast cancer onset 
         or end of follow-up period",
         subtitle = "Breast cancer tended to develop sooner for patients with
         radiotherapy than those without.")
```

### Breast cancer incidents

```{r}
df = dat %>% mutate(STATUS = ifelse(STATUS == 1, "Breast cancer", 
                                    "No breast cancer"))
plt = mk_barplot_freq(df)
p = plt("RT", "STATUS")
add_labs(p, title = "Distribution of event status by treatment",
         subtitle = "Breast cancer incidence among patients without radiotherapy
         is more than double of that among patients with radiotherapy.")
```


### Kaplan Meier

```{r}
km = survfit(Surv(TTOBRC, STATUS) ~ RT, data = dat, type = "kaplan-meier")        
plot(km, lty=c(1, 1), lwd=c(3, 3), col=c(red, blue), ylim=c(0.86, 1), 
     xlab = "Months to breast cancer onset or end of follow-up period", 
     ylab = "Proportion of cancer-free patients", 
     main = "Kaplan-Meier", 
     cex.axis = 1, cex.lab = 1, cex.main=1.5)
legend("topright", lty=c(1, 1), lwd=c(3, 3), text.font=1.5,
       legend = c("KM: no radiotherapy", "KM: radiotherapy"), 
       col = c(red, blue))
```

***

Looking at the Kaplan-Meier curves on the left, we see in the short term, 
patients who received radiotherapy had a higher proportion of cancer-free 
patients and hence lower breast cancer incidence than patients without 
radiotherapy, but in the long term, patients with radiotherapy had a higher 
breast cancer incidence than patients without. This observation leads to the 
following hypothesis: radiotherapy decreases the risk of breast cancer in the 
short term, but increases the risk in the long term.


### Cox Model

```{r}
# run cox model
cox = coxph(Surv(TTOBRC, STATUS) ~ RT, data=dat, method="breslow")
df_coef = broom::tidy(cox) %>% 
        mutate(`exp(est)` = exp(estimate)) %>% 
        rename(est = estimate, `p-val` = p.value) %>% 
        select(term, est, `exp(est)`, `p-val`)

# calculate predicted survival rates
survcurve.cox.notreat = survexp(~RT, data=radio, ratetable=cox, cohort=TRUE)
survcurve.cox.treat = survexp(~RT, data=noradio, ratetable=cox,  cohort=TRUE)

# generate KM plot with survival curves overlayed 
km = survfit(Surv(TTOBRC, STATUS) ~ RT, data=dat, type="kaplan-meier")
plot(km, lty=c(1, 1), lwd=c(3, 3), col=c(red, blue), ylim=c(0.86, 1), 
     xlab = "Months to breast cancer onset or end of follow-up period", 
     ylab = "Proportion of cancer-free patients", 
     main = "Kaplan-Meier vs. Cox Model",
     cex.axis = 1, cex.lab = 1, cex.main=1.5)

lines(survcurve.cox.notreat, col=green, ylim=c(0.86,1), lty=1, lwd=3)
lines(survcurve.cox.treat, col=green, ylim=c(0.86,1), lty=2, lwd=3)
legend("topright", lty=c(1,1,1,2), lwd=rep(3, 4), text.font=1.5,
       legend=c("KM: no radiotherapy", "KM: radiotherapy", 
                "Cox: no radiotherapy", "Cox: radiotherapy"), 
       col=c(red, blue, green, green))
```

***
```{r}
knitr::kable(df_coef, digits = 3, caption = "Model parameter estimate")
```

The parameter estimate of the treatment variable has a p-value greater than 0.05, indicating the effect of radiotherapy is not statistically significant under 
the Cox model.

This plot on the left shows the Cox model fits the data poorly. The cox model 
fails to describe the data because it doesn't account for the possibility that
short and long term effects of the treatment on the hazard can be in opposite direction.

### PO Model

```{r}
# download source from https://cran.r-project.org/src/contrib/Archive/nltm/
# install pkg via
# install.packages("~/Coding/R/Rpkg/nltm", repos=NULL, type="source")

# run po model
invisible(capture.output(
        po <- nltm::nltm(Surv(TTOBRC, STATUS)~RT, data=dat, nlt.model="PO")
        ))

# extract beta hats and collect into the correct format for output
summ = summary(po)
vars = c("coef", "exp(coef)", "p")
df_coef = c(row.names(summ$coef), round(summ$coef[, vars], 3))
df_coef = data.frame(rbind(df_coef))
names(df_coef) = c("term", "est", "exp(est)", "p-val")
row.names(df_coef) = NULL

# calculate survival probabilities for (no radiotherapy) and (radiotherapy)
po.baselineSF = cumprod(po$surv)
po.noradioSF = 1/(1-log(po.baselineSF))
betas = po$coef
po.radioSF = exp(betas)/(exp(betas)-log(po.baselineSF))
eventTimes = sort(unique(dat$TTOBRC[dat$STATUS==1]))        

# generate KM plot with survival curves overlayed 
km = survfit(Surv(TTOBRC, STATUS) ~ RT, data=dat, type="kaplan-meier")        
plot(km, lty=c(1, 1), lwd=c(3, 3), col=c(red, blue), ylim=c(0.86, 1), 
     xlab = "Months to breast cancer onset or end of follow-up period", 
     ylab = "Proportion of cancer-free patients", 
     main = "Kaplan-Meier vs. PO Model",
     cex.axis = 1, cex.lab = 1, cex.main=1.5)

lines(eventTimes, po.noradioSF, type='s', col=yellow, lty=1, lwd=3, ylim=c(0.86,1))
lines(eventTimes, po.radioSF, type='s', col=yellow, lty=2, lwd=3, ylim=c(0.86,1))
legend("topright", legend=c("KM: no radiotherapy", "KM: radiotherapy", 
                            "PO: no radiotherapy", "PO: radiotherapy"),
       lty=c(1, 1, 1, 2), lwd=rep(3, 4), text.font=1.5, 
       col=c(red, blue, yellow, yellow)) 
```

***
```{r}
knitr::kable(df_coef, digits = 3, caption = "Model parameter estimate")
```

The parameter estimate of the treatment variable has a p-value greater than 0.05, indicating the effect of radiotherapy is not statistically significant under the proportional odds model.

The plot on the left shows the proportional odds model fits the data poorly. The
PO model fails to describe the data because the it also doesn't account for the 
possibility that short and long term effects of the treatment on the hazard can 
be in opposite direction.



### PHPH Model

```{r}
# run phph model
invisible(capture.output(
        phph <- nltm::nltm(Surv(TTOBRC, STATUS)~RT, data=dat, nlt.model="PHPHC")
        ))

# extract beta hats and collect into the correct format for output
summ = summary(phph)
vars = c("coef", "exp(coef)", "p")
row1 = c(row.names(summ$coef)[1], round(summ$coef[1, vars], 3))
row2 = c(row.names(summ$coef)[2], round(summ$coef[2, vars], 3))
row3 = c(row.names(summ$coef)[3], round(summ$coef[3, vars], 3))
df_coef = data.frame(rbind(row1, row2, row3))
names(df_coef) = c("term", "est", "exp(est)", "p-val")
row.names(df_coef) = NULL

# calculate survival probabilities for (no radiotherapy) and (radiotherapy)
CureRate = phph$coeff[length(phph$coeff)]
betas = phph$coeff
# long-term predictor for subjects with radiotherapy treatment
betasCure = betas[1:phph$nvar$pred.long]
logtheta = CureRate + betasCure
theta = exp(logtheta)
# Short-term predictor for subjects with radiotherapy treatment
betasNonCured = betas[phph$nvar$pred.long + 1:phph$nvar$pred.short]
logeta = betasNonCured
eta = exp(logeta)
# unique event times sorted
eventTimes = sort(unique(dat$TTOBRC[dat$STATUS==1]))
# baseline survival function values corresponding to eventTimes
phph.baselineSF = cumprod(phph$surv)
# Survival function for subjects with radiotherapy
phph.radioSF = exp(-theta*(1-phph.baselineSF^eta))
# Survival function for subjects without radiotherapy
phph.noradioSF = exp(-exp(CureRate)*(1-phph.baselineSF))

# generate KM plot with survival curves overlayed 
km = survfit(Surv(TTOBRC, STATUS) ~ RT, data=dat, type="kaplan-meier")        
plot(km, lty=c(1, 1), lwd=c(3,3), col=c(red, blue), ylim=c(0.86, 1), 
     xlab = "Months to breast cancer onset or end of follow-up period", 
     ylab = "Proportion of cancer-free patients", 
     main = "Kaplan-Meier vs. PHPH Cure Model",
     cex.axis = 1, cex.lab = 1, cex.main=1.5)

lines(eventTimes, phph.noradioSF, type='s', lty=1, lwd=3, ylim=c(0.86,1), col=purple)
lines(eventTimes ,phph.radioSF, type='s', lty=2, lwd=3, ylim=c(0.86,1), col=purple)
legend("topright", legend=c("KM: no radiotherapy", "KM: radiotherapy", 
                            "PHPH: no radiotherapy", "PHPH: radiotherapy"),
       lty=c(1,1,1,2), lwd=rep(3,4), text.font=2, 
       col=c(red, blue, purple, purple))         
```


***
```{r}
knitr::kable(df_coef, digits = 3, caption = "Model parameter estimate")
```

The p-values are tiny, indicating the beta estimates are all statistically significant. In particular, the significant beta estimate for radiotherapy in the short term predictor implies radiotherapy has a negative effect on breast cancer
incidence in the short term; the significant beta estimate for radiotherapy in 
the long term predictor implies radiotherapy has a positive effect on breast 
cancer incidence in the long run.

The plot on the left shows the PHPHC cure model fits the data well. This is 
because the phph cure model includes a term that captures the short term effect explicitly.

